\chapter{Evaluation}

In the preparation chapter, the four modified success criteria were laid out:

\begin{enumerate}
    \item The tool should be suitable for a complete beginner to use, avoiding unexplained technical language, and starting with challenges of an appropriate level.

    \item The tool should allow full flexibility in which features are used; turning all additional features off should make it function as a simple code editor. Additional features should include the ability to restrict programming constructs and visualise programs.

    \item The tool’s features should be useful for both complete beginners as well as those with some small experience in Python. At least three challenges should be provided for any user learning beginner-level features (looping, control flow, basic data structures, etc.)

    \item I will test the run-time overhead added by using the tool. The tool should not add significant overhead, so that the programs run are still usable.
\end{enumerate}

The first three were evaluated as part of a usability study, described in section 4.1. Criterion 4 was evaluated using timing analysis, which is discussed in section 4.2.

\section{Study}

This section details the study performed to evaluate the app. First, the research questions are discussed, followed by description of the study design.

\subsection{Research Questions}

With the set of success criteria in mind, the following research questions were established to evaluate the overall usability and efficacy of the app.

\subsubsection{Efficacy}
\textbf{RQ 1: }To what extent does the tool help users to learn coding? 

This RQ aims to evaluate the tool's efficacy, in particular by assessing whether users perceive that the tool helps them learn.

\textbf{RQ 2: }Are the challenges provided by the tool appropriately scoped for beginner learners?

This RQ aims to directly address criterion 3 by assessing whether the tool is appropriately scoped to beginners.

\subsubsection{Usability}
\textbf{RQ 3: }How usable is the tool for beginners?

This RQ aims to address criterion 1 by assessing whether beginners were able to understand how to use the tool without help.

\textbf{RQ 4: }How usable is the tool as a simple code editor?

This RQ aims to address criterion 2 by assessing whether the tool can transition with users as they learn, eventually leaving them competent with a standard code editor.

\subsection{Study Design}

The questionnaire consisted of two main sections. The chosen questions needed to accurately measure what they claimed. For the usability section, the System Usability Scale \cite{brooke_sus_1995} was used. It was suitable because...

\subsection{Results}


\subsection{Discussion}

\section{Timing stuff}