\chapter{Introduction}

In this project, a Python programming environment is created targeted at new programmers. The environment builds upon the idea of helping users learn by removing some of the complexity behind programming, creating a less overwhelming experience. In particular, this project focuses on restricting the expressiveness of Python, then gradually increasing it in line with the user's knowledge. This is done by creating an environment that restricts both language features and built-in functions.

*WHEN WRITING CONCLUSION ADD SOMETHING ABOUT HOW WELL THE PROJECT DOES THIS HERE*

\section{Motivation}

One problem many novice programmers face is the overwhelming nature of programming. The combination of learning many abstract concepts and a completely new syntax for representing ideas leads to difficulty \cite{green_problems_1980,lahtinen_study_2005}. Creating new tools to help them overcome these challenges has been a focus of computing education research. The idea behind many tools is to create a more rigid programming environment, removing some aspect of complexity from the language \cite{bau_learnable_2017, hermans_hedy_2020, valente_simplifying_2022}.
Students often struggle the most with understanding exactly how a computer takes code and reaches the output \cite{lahtinen_study_2005}. Notional machines help explain why; a notional machine is a simplified mental model of a computer which helps learners think through how a program is executed \cite{sorva_notional_2013}. When users face a large, daunting syntax, there is a mismatch between their notional machine and the language's full capabilities. Thus, difficulty emerges in explaining behaviour, as the user's concept of how the computer works does not match reality \cite{hermans_hedy_2020}.
Adding restrictions reduces this discrepancy.

Block-based languages, such as Scratch\footnote{\href{https://scratch.mit.edu/}{https://scratch.mit.edu/}} or Blockly\footnote{\href{https://developers.google.com/blockly}{https://developers.google.com/blockly}} have seen success in recent years- as of 2024, Scratch had 140 million users \cite{noauthor_2024_2024}. This begs the question of why a restrictive, non-block-based approach could be useful. There are several reasons for this.

First, learners eventually need to transition from block-based to text-based to continue progressing, as there are several advantages for users. Text-based languages provide a host of useful libraries that greatly improve the experience of writing complex code and improve expressiveness. Additionally, for experienced users, it will likely be quicker to type than searching menus and dragging code blocks, and they may want to write more efficient code, which generally requires a lower-level language. \citet{kazemitabaar_scaffolding_2023} found that a structured middle ground --- as would be provided by this project --- helps reduce errors that learners make while transitioning.

Moreover, having a variety of approaches ensures a range of learners are accounted for. People learn in different ways; some may find the block-based approach too slow to use \cite{weintrop_block_2015}, for example. \citet{weintrop_block_2015} found that some older learners perceived block-based languages to not be `real' programming. These people may prefer a tool based on an established programming language, such as Python. Furthermore, Python has extensive support through libraries. A user wanting to learn to utilise a specific library may find it quicker to learn directly in Python. Many school curricula, including those used in England, require learning text-based programming languages, but not block-based ones \cite{noauthor_national_2013}. Consequently, many students who have not experienced block-based languages are forced to learn text-based languages. Thus, a text-based learning environment is useful. Finally, these approaches are not mutually exclusive. Testing this approach on a text-based language may point towards success when the same restrictive approach is applied to a block-based language.

Another benefit of the restriction approach comes from the ability to remove well-understood features to prompt users towards new features. For example, a student who is familiar with looping might try to apply them to solve problems when being taught recursion. Reapplying known ideas like this is an invaluable part of programming \cite{soloway_empirical_1984}, but is one novices often do incorrectly \cite{robins_learning_2003}; novices often lack an understanding of why code works \cite{lister_multi-national_2005}. By removing such features, the student is forced out of their comfort zone --- a key aspect of learning \cite{chaiklin_zone_2003} --- and encouraged to gain an understanding of the problem.

The tool also uses program visualisation, a concept employed by several learning platforms. Restrictions are incorporated into the design of this tool's visualiser; its primary purpose is to help learners understand how restrictions affect the execution of their program. Thus helping users update their notional machine and better understand execution: a key area of difficulty for learners \cite{lahtinen_study_2005}. \citet{sorva_review_2013} identifies several reasons educators ignore visualisations. These include poor usability or doubts about effectiveness. By creating a fully integrated and usable visualiser and taking a unique approach to visualisation, the project aims to remove these concerns.

*THIS PARAGRAPH NEEDS REWRITING AND EXPANDING, SEE LAURIE'S COMMENTS*

\section{Definitions}
The following definitions are used in the rest of the report.

A \textit{restriction} consists of removing syntax from the language; code containing restrictions cannot be run, just as code containing a syntax error cannot. Restrictions come in two forms: 
\begin{itemize}
    \item \textit{Language features}, henceforth \textit{features}: 
    Programming language constructs and data structures (e.g. loops, conditionals, functions, lists).
    \item \textit{Built-in functions}, henceforth \textit{built-ins}: The set of functions defined on the ``Built-in Functions'' page of the Python documentation\footnote{\href{https://docs.python.org/3/library/functions.html}{https://docs.python.org/3/library/functions.html}}.
\end{itemize}

Establishing these definitions was one of the first pieces of the preparation work.


**IN HERE MENTION AND CITE SOMETHING ABOUT HOW CREATING RESTRICTIONS HELPS USERS TO LEARN NEW CONCEPTS WITHOUT RELYING ON OLD ONES.